---
title: "Chapter_3_TS-decomposition"
format: 
  html:
    code-fold: true
    toc: true
    toc-location: left
execute:
  warning: false
editor: visual
---

## STL decomposition. 23 March 2025.

```{r}
#| echo: true

library(fpp2)
library(fpp3)
```

### Data: US retail employment

```{r}

us_employment  
```

### Example 3.2

```{r}

us_retail_employment <- us_employment |>
  filter(Title == "Retail Trade") |>
  filter(year(Month)>=1990) |>
  select(-Series_ID)

us_retail_employment

```

```{r}
autoplot(us_retail_employment, Employed)+
  labs(y="Persons (thousands)",
       title = "Increasing trend of employment in the US retail-sector")
```

#### STL decomposition

```{r}
dcmp <- us_retail_employment |>
  model(stl = STL(Employed))

dcmp
```

```{r}
components(dcmp)
```

This object is called a dable, a decomposition-table, which provides details about the following:

-   model

-   time-period for which the data was recorded, in this case "Month" (yearmonth)

-   the actual value for the data "Employed"

-   decomposed parts = trend, seasonality, remainder, seasonally_adjusted (Employed- season_year)

#### Plotting trends from the dable

```{r}

components(dcmp) |>
  as_tsibble() |>
  autoplot(Employed, colour="gray")+
  geom_line(aes(y=trend), colour="orange")+
  labs(
    y="Persons (thousands)",
    title = "Trend for the US retail employment 1990-2020"
  )

```

#### Plot all compononents

```{r}
components(dcmp) |> autoplot()
```

## STL function: Section 3.6

```{r}
us_retail_employment |>
  model(
    STL(Employed ~ trend(window=7)+
          season(window = "periodic"),
        robust = TRUE)
  ) |>
  components() |>
  autoplot()
```

Observations:

-   The STL model had two parameters:

    -   trend window, which basically accounts for the number of consecutive observations we should use to calculate the trend. The higher the trend-window the smoother the curve.

    -   season-window, gives the number of years that shaould be taken to calculate the seasonal-values. When this is set to infinite, which implies that the seasonal-trend remains constant through the data-period, we use the term "periodic".

    -   Defualt settings in the package:

        -   default for season = 11

        -   defalut for trend varies with the seoasnality.

            -   for monthly data, trend default = 21

    -   Read the package for more details

## Exercises:
